A weakly structured stem for human origins in Africa

### 2022-03-11 paper.tex
real-time edits

\comment{About title: "Contrasting archaic admixture and a weakly structured stem for human originas in Africa" Too long? Inconclusive?} \sgcomment{Disagree.}

\begin{abstract}
While it is now broadly accepted that \emph{Homo sapiens}
origined within Africa, considerable
uncertainty surrounds specific models of divergence and migration across the
continent. Progress is hampered by a paucity of fossil and genomic data, as
well as variability in prior dating estimates.
Here we use linkage disequilibrium and
diversity-based statistics, optimized for rapid, complex demographic inference
to discriminate among such models \comment{\ldots to evaluate the fit of various models}.
We infer detailed demographic models for
populations across Africa, including representatives from eastern and western
groups, as well as 44 newly whole-genome sequenced individuals from the Nama
(Khoe-San). Despite the complexity of African population history, contemporary
population structure dates back to Marine Isotope Stage (MIS) 5. The earliest
population divergence among contemporary populations occurs 120-135ka, between
the Khoe-San and other groups. Prior to the divergence of contemporary African
groups, we infer long-lasting structure between two or more weakly
differentiated ancestral \emph{Homo spp.}
\comment{maybe don't abbreviate; won't be understood outside of evolutionary biology as much}
populations connected by gene flow over
hundreds of thousands of years (i.e. a weakly structured stem). We find that
weakly structured stem models provide more likely explanations of polymorphism
that had previously been attributed to contributions from archaic hominins in
Africa. \aprcomment{.. to contributions from archaic hominin introgression in Africa?}
In contrast to models with archaic introgression, we predict that fossil
remains from coexisting ancestral populations should be morphologically
similar.
Despite genetic similarity between these populations, an inferred 1--4\%
\sgcomment{explain in the main text where this estimate is from!}
of genetic differentiation among contemporary human populations can be
attributed to genetic drift between stem populations.
We show that model misspecification explains variation in previous
divergence time estimates and argue that studying a suite of models is key to
robust inferences about deep history.
\end{abstract}

@@

Archaeological sites from the Middle Stone Age (approx. 300ka-40ka) are widely
distributed across Africa, and are particularly well represented in the
northern, eastern and southern parts of the continent. Similarly, fossil crania
such as those from the sites of Jebel Irhoud \citep{Hublin2017-cq}, Herto
\citep{White2003-bk} and Klasies River \citep{Deacon1995-rx} demonstrate that
anatomically derived \emph{Homo sapiens} features were also present across the
continent during this period. It has been difficult to reconcile these lines of
evidence with evidence from genomics, which have suggested a predominantly
tree-like model of recent population divergence from a single ancestral
population. It is unclear whether fossil specimens and archaeological sites
represent populations which contributed to our ancestors as population
precedents, or were local ``dead-ends'' from which contemporary \emph{Homo
sapiens} do not descend. 
\comment{unpack more; explain why these are at odds}
Recently, synthetic \comment{synthetic doesn't feel quite right here, can delete}
attempts to reconcile genetic and
paleoanthropological data include proposals for a Pan-African origin of
\emph{Homo sapiens} by which populations in many regions of the continent
contributed to the formation of Homo sapiens beginning at least 300ka
\citep{Stringer2016-mj,Scerri2018-nl,Scerri2019-xg}.

Genetic models have been hampered in their contribution to this discussion
because they primarily assume (or, at least, have been tested under) a
tree-like model of isolation-with-migration. Alternative theoretical scenarios
have been proposed, such as stepping stone models \citep{Arredondo2021-qa} or
population coalescence and fragmentation \citep{Scerri2019-xg}. These
approaches are more challenging to interpret and fit to data.
\comment{I think a schematic illustrating models being tested would be extremely useful in the main text. This could be Fig S2, or even a new figure with a further simplified illustration o fthe general idea of the models being tested to help orient the reader.}
However, new
population genetic tools now allow for inference involving tens to
hundreds of genomes from multiple populations and greater complexity
\citep{Kamm2020-vn,Ragsdale2019-nt,Speidel2019-nj}. Inspired by evidence for
Neanderthal admixture with humans in Eurasia, several recent articles
have shown that introducing an archaic hominin ghost population contributing to African
populations in the period surrounding the Out-of-Africa migration event
substantially improves the description of genetic data relative to
single-origin models
\citep{Plagnol2006-lt,Hammer2011-bx,Hsieh2016-gk,Hey2018-pw,Ragsdale2019-nt,Durvasula2020-td,Lorente-Galdos2019-vz,Durvasula2020-td}.
This has driven speculation about the geographic range of this ghost
population, possible links to specific fossils, and the possibility of
finding ancient DNA evidence \citep{Hsieh2016-gk}). However, these prior
articles share two weaknesses. First, they only contrast a single-origin model
with an archaic hominin admixture model, leaving out other plausible models
\citep{Henn2018-rf} (Figure~\ref{fig:supp-possible-models}).
Second, they
focus on a small subset of African diversity, either because of small sample
sizes (2-5 genomes) or because they rely on 1000 Genomes data which only
recruited populations of recent West African ancestry (Figure \ref{fig:1}C). 
\comment{What about the LWK?}
While ancient DNA from Eurasia has helped us understand early human history
outside of Africa, there is no comparably ancient DNA to elucidate early history
in Africa \citep{Lipson2022-xf}.

Here, we therefore aim to discriminate among a broader set of demographic
models by interrogating the genomes of contemporary populations. We take as our
starting point 4 classes of models (single population expansion, single
population expansion with regional persistence, archaic hominin admixture, and
multi-regional evolution, Figure~\ref{fig:supp-possible-models}), using 290
genomes from southern, eastern, and western Africa as well as Eurasia. By
including geographically and genetically diverse populations across Africa, we
infer demographic models that explain more aspects of genetic diversity
\comment{this is a bit vague; make more detailed. e.g.: better fit patterns of LD and \ldots}
in more
populations than previously reported. These analyses confirm the inadequacy of
tree-like models and provide an opportunity to directly evaluate a wide range
of alternative models.

@@

We inferred detailed demographic histories using 4x-8x whole-genome sequencing
data for four diverse African populations, comprising the Nama (Khoe-San from
South Africa, newly presented here), Mende (from Sierre Leone, MSL from the
Phase 3 1000 Genomes Project \citep{1000_Genomes_Project_Consortium2015-zq}),
Gumuz (recent descendents of a hunter-gatherer group from Ethiopia
\citep{Gurdasani2015-qy,Gopalan2019-wd}), and Eastern African agriculturalists
(Amhara and Oromo from Ethiopia \citep{Gurdasani2015-qy}). The Amhara and Oromo
populations, despite speaking distinct Afro-Asiatic languages, are highly
genetically similar \citep{Pagani2015-pz,Gopalan2019-wd} and thus the two
groups were combined for a larger sample size (Figure~\ref{fig:1}). We also
included the British (GBR) from the 1000 Genomes Project in our demographic
models as a representative source of back-to-Africa gene flow and recent
colonial admixture in South Africa. Finally, we used a high-coverage ancient
Neanderthal genome from Vindija Cave, Croatia \citep{Prufer2017-kk} to account
for gene flow from Neanderthals into non-Africans and gauge the relative
time depth of divergence, assuming Neanderthals diverged 550ka from a common
stem. For each population, we computed low-order
\comment{what do you mean by low-order? not a commonly used phrase}
\aprcomment{I think this is a fairly common term in statistics, right?}
allele frequency and linkage
disequilibrium (LD) statistics that are well suited for both low- and
high-coverage genomes \sgcomment{We computed one- and two-locus statistics whose expectation within and across populations can be computed efficiently and that are well suited for both low- and
high-coverage genomes} \citep{Ragsdale2019-nt,Ragsdale2020-nz}. Using a
maximum-likelihood inference framework, we then fit to these statistics a
family of parameterized demographic models that involve population splits, size
changes, continuous and variable migration rates, and punctuated admixture
events, to learn about the nature of population structure over the past million
years.

\begin{figure}[ht!]
    \centering
    \includegraphics{figures/fig1.pdf}
    \caption[width=\textwidth]{
        \textbf{Genetic diversity across Africa.}
        (A) Select populations from the 1000 Genomes and African Diversity Reference Panels
        illustrate diversity from western, eastern and southern Africa.
        We chose representatives from each region (bold labels)
        to build parameterized models,
        including the newly-sequenced Nama from South Africa, Mende
        from Sierre Leone, Gumuz, Oromo and Amhara from Ethiopia, and the
        British and Vindija Neanderthal individual.
        (B, C) PCA highlights the range of genetic divergence anchored 
        by western Africans, Nama, Gumuz and the British.
        Percentages show variance explained by each principal component.
        (D) \texttt{ADMIXTURE} illustrates signatures of recent gene flow in Africa
        which reflect colonial-period migration into the Nama,
        Back-to-Africa gene flow among some Ethiopians, and Khoe-San admixture in the Zulu.
        \comment{How did you land on the best fit k? Show scree plot of CV errors, or otherwise justify choice}
    }
    \label{fig:1}
\end{figure}

@@

A population reticulation involves multiple stems contributing genetically to
the formation of a group. One way in which this can happen is through the
geographic expansion of one or both stems. For example, if during MIS 5, either
Stem 1S \comment{where is this illustrated/labeled?}
from southern Africa moved northward thereby encountering the Stem 2,
or Stem 2 moved from central/western Africa southward into Stem 1S -- then we
could observe disproportionate ancestry contributions from different stems in
contemporary groups. We observed two merger events. The first, between Stem 1S and
Stem 2, results in the formation of an ancestral Khoe-San population ~120ka.
The second $\approx100$ka between Stem 1E and Stem 2, results in the formation of the
ancestors of East/West Africans as well as later ``Out of Africans''. The rapid
rise in sea levels and increased precipitation during MIS 5e, following a
glacial period of aridity across Africa \citep{Blome2012-lw}, might have
triggered migration inland away from the coasts, as has been suggested for the
Paleo-Agulhas plain \citep{Marean2014-pg}.
\comment{very specific... might be worth expanding a bit}

@@

There is evidence for both deleterious and adaptive archaic-hominin-derived alleles in
contemporary genomes in the form of a depletion of Neanderthal ancestry in regulatory
regions \citep{Petr2019-xo} or an increased frequency of archaic-hominin-related
haplotypes such as at \emph{EPAS1} among Tibetans \citep{Zhang2021-xx}. Under previous
African archaic hominin admixture models, the estimated 8--10\% introgression rate is
much higher than Neanderthal gene flow, and would have plausibly been fertile
ground for dramatic selection for or against archaic-hominin-derived haplotypes\citep{Wall2019-ao}. By contrast, adaptation under a weakly
structured stem would have occurred continuously over much longer periods.
Patterns of polymorphism that are inconsistent with the single-stem model
predictions have been used to infer putative archaic admixed segments
\citep{Plagnol2006-lt,Hsieh2016-gk,Wall2019-ao,Durvasula2020-td}, negative
selection against such segments \citep{Wall2019-ao}, and pervasive positive
selection \citep{Schrider2017-kl}. However, such approaches are subject to high
false positives in the presence of population structure with migration
\citep{Petr2019-xo}, and their interpretation should be re-examined in light of
a weakly-structured-stem model within Africa. \comment{Reviewers may suggest this demonstrated here}

>>>>

\begin{abstract}
While it is now broadly accepted that \emph{Homo sapiens}
origined within Africa, considerable
uncertainty surrounds specific models of divergence and migration across the
continent. Progress is hampered by a paucity of fossil and genomic data, as
well as variability in prior dating estimates.
Here we use linkage disequilibrium and
diversity-based statistics, optimized for rapid, complex demographic inference
to discriminate among such models \comment{\ldots to evaluate the fit of various models}.
We infer detailed demographic models for
populations across Africa, including representatives from eastern and western
groups, as well as 44 newly whole-genome sequenced individuals from the Nama
(Khoe-San). Despite the complexity of African population history, contemporary
population structure dates back to Marine Isotope Stage (MIS) 5. The earliest
population divergence among contemporary populations occurs 120-135ka, between
the Khoe-San and other groups. Prior to the divergence of contemporary African
groups, we infer long-lasting structure between two or more weakly
differentiated ancestral \emph{Homo}
populations connected by gene flow over
hundreds of thousands of years (i.e. a weakly structured stem). We find that
weakly structured stem models provide more likely explanations of polymorphism
that had previously been attributed to contributions from archaic hominins in
Africa.
In contrast to models with archaic introgression, we predict that fossil
remains from coexisting ancestral populations should be morphologically
similar.
Despite genetic similarity between these populations, an inferred 1--4\%
\sgcomment{explain in the main text where this estimate is from!}
of genetic differentiation among contemporary human populations can be
attributed to genetic drift between stem populations.
We show that model misspecification explains variation in previous
divergence time estimates and argue that studying a suite of models is key to
robust inferences about deep history.
\end{abstract}

@@

Archaeological sites from the Middle Stone Age (approx. 300ka-40ka) are widely
distributed across Africa, and are particularly well represented in the
northern, eastern and southern parts of the continent. Similarly, fossil crania
such as those from the sites of Jebel Irhoud \citep{Hublin2017-cq}, Herto
\citep{White2003-bk} and Klasies River \citep{Deacon1995-rx} demonstrate that
anatomically derived \emph{Homo sapiens} features were also present across the
continent during this period. It has been difficult to reconcile these lines of
evidence with evidence from genomics, which have suggested a predominantly
tree-like model of recent population divergence from a single ancestral
population. It is unclear whether fossil specimens and archaeological sites
represent populations which contributed to our ancestors as population
precedents, or were local ``dead-ends'' from which contemporary \emph{Homo
sapiens} do not descend. 
Recently,
attempts to reconcile genetic and
paleoanthropological data include proposals for a Pan-African origin of
\emph{Homo sapiens} by which populations in many regions of the continent
contributed to the formation of Homo sapiens beginning at least 300ka
\citep{Stringer2016-mj,Scerri2018-nl,Scerri2019-xg}.

Genetic models have been hampered in their contribution to this discussion
because they primarily assume (or, at least, have been tested under) a
tree-like model of isolation-with-migration. Alternative theoretical scenarios
have been proposed, such as stepping stone models \citep{Arredondo2021-qa} or
population coalescence and fragmentation \citep{Scerri2019-xg}. These
approaches are more challenging to interpret and fit to data.
\comment{I think a schematic illustrating models being tested would be extremely useful in the main text. This could be Fig S2, or even a new figure with a further simplified illustration o fthe general idea of the models being tested to help orient the reader.}
However, new
population genetic tools now allow for inference involving tens to
hundreds of genomes from multiple populations and greater complexity
\citep{Kamm2020-vn,Ragsdale2019-nt,Speidel2019-nj}. Inspired by evidence for
Neanderthal admixture with humans in Eurasia, several recent articles
have shown that introducing an archaic hominin ghost population contributing to African
populations in the period surrounding the Out-of-Africa migration event
substantially improves the description of genetic data relative to
single-origin models
\citep{Plagnol2006-lt,Hammer2011-bx,Hsieh2016-gk,Hey2018-pw,Ragsdale2019-nt,Durvasula2020-td,Lorente-Galdos2019-vz,Durvasula2020-td}.
This has driven speculation about the geographic range of this ghost
population, possible links to specific fossils, and the possibility of
finding ancient DNA evidence \citep{Hsieh2016-gk}). However, these prior
articles share two weaknesses. First, they only contrast a single-origin model
with an archaic hominin admixture model, leaving out other plausible models
\citep{Henn2018-rf} (Figure~\ref{fig:supp-possible-models}).
Second, they
focus on a small subset of African diversity, either because of small sample
sizes (2-5 genomes) or because they rely on 1000 Genomes data which only
recruited populations of recent West African or Bantu-speaking ancestry (Figure \ref{fig:1}C). 
While ancient DNA from Eurasia has helped us understand early human history
outside of Africa, there is no comparably ancient DNA to elucidate early history
in Africa \citep{Lipson2022-xf}.

Here, we therefore aim to discriminate among a broader set of demographic
models by interrogating the genomes of contemporary populations. We take as our
starting point 4 classes of models (single population expansion, single
population expansion with regional persistence, archaic hominin admixture, and
multi-regional evolution, Figure~\ref{fig:supp-possible-models}), using 290
genomes from southern, eastern, and western Africa as well as Eurasia. By
including geographically and genetically diverse populations across Africa, we
infer demographic models that explain more features of genetic diversity
in more
populations than previously reported. These analyses confirm the inadequacy of
tree-like models and provide an opportunity to directly evaluate a wide range
of alternative models.

@@

We inferred detailed demographic histories using 4x-8x whole-genome sequencing
data for four diverse African populations, comprising the Nama (Khoe-San from
South Africa, newly presented here), Mende (from Sierre Leone, MSL from the
Phase 3 1000 Genomes Project \citep{1000_Genomes_Project_Consortium2015-zq}),
Gumuz (recent descendents of a hunter-gatherer group from Ethiopia
\citep{Gurdasani2015-qy,Gopalan2019-wd}), and Eastern African agriculturalists
(Amhara and Oromo from Ethiopia \citep{Gurdasani2015-qy}). The Amhara and Oromo
populations, despite speaking distinct Afro-Asiatic languages, are highly
genetically similar \citep{Pagani2015-pz,Gopalan2019-wd} and thus the two
groups were combined for a larger sample size (Figure~\ref{fig:1}). We also
included the British (GBR) from the 1000 Genomes Project in our demographic
models as a representative source of back-to-Africa gene flow and recent
colonial admixture in South Africa. Finally, we used a high-coverage ancient
Neanderthal genome from Vindija Cave, Croatia \citep{Prufer2017-kk} to account
for gene flow from Neanderthals into non-Africans and gauge the relative
time depth of divergence, assuming Neanderthals diverged 550ka from a common
stem. We computed one- and two-locus statistics whose expectation within and across 
populations can be computed efficiently and that are well suited for both low- and
high-coverage genomes \citep{Ragsdale2019-nt,Ragsdale2020-nz}. Using a
maximum-likelihood inference framework, we then fit to these statistics a
family of parameterized demographic models that involve population splits, size
changes, continuous and variable migration rates, and punctuated admixture
events, to learn about the nature of population structure over the past million
years. \sgcomment{define LD}

\begin{figure}[ht!]
    \centering
    \includegraphics{figures/fig1.pdf}
    \caption[width=\textwidth]{
        \textbf{Genetic diversity across Africa.}
        (A) \sgcomment{secure coordinates} Select populations from the 1000 Genomes and African Diversity Reference Panels
        illustrate diversity from western, eastern and southern Africa.
        We chose representatives from each region (bold labels)
        to build parameterized models,
        including the newly-sequenced Nama from South Africa, Mende
        from Sierre Leone, Gumuz, Oromo and Amhara from Ethiopia, and the
        British and Vindija Neanderthal individual.
        (B, C) PCA highlights the range of genetic divergence anchored 
        by western Africans, Nama, Gumuz and the British.
        Percentages show variance explained by each principal component.
        (D) \texttt{ADMIXTURE} with $K=4$ illustrates signatures of recent gene flow in Africa
        which reflect colonial-period migration into the Nama,
        Back-to-Africa gene flow among some Ethiopians, and Khoe-San admixture in the Zulu.
        %\comment{How did you land on the best fit k? Show scree plot of CV errors, or otherwise justify choice}
    }
    \label{fig:1}
\end{figure}

@@

A population reticulation involves multiple stems contributing genetically to
the formation of a group. One way in which this can happen is through the
geographic expansion of one or both stems. For example, if during MIS 5, either
Stem 1S (Figure \ref{fig:2}B) 
from southern Africa moved northward thereby encountering the Stem 2,
or Stem 2 moved from central/western Africa southward into Stem 1S -- then we
could observe disproportionate ancestry contributions from different stems in
contemporary groups. We observed two merger events. The first, between Stem 1S and
Stem 2, results in the formation of an ancestral Khoe-San population ~120ka.
The second $\approx100$ka between Stem 1E and Stem 2, results in the formation of the
ancestors of East/West Africans as well as later ``Out of Africans''. The rapid
rise in sea levels and increased precipitation during MIS 5e, following a
glacial period of aridity across Africa \citep{Blome2012-lw}, might have
triggered migration inland away from the coasts, as has been suggested, e.g., for the
Paleo-Agulhas plain \citep{Marean2014-pg}.

@@

There is evidence for both deleterious and adaptive archaic-hominin-derived alleles in
contemporary genomes in the form of a depletion of Neanderthal ancestry in regulatory
regions \citep{Petr2019-xo} or an increased frequency of archaic-hominin-related
haplotypes such as at \emph{EPAS1} among Tibetans \citep{Zhang2021-xx}. Under previous
African archaic hominin admixture models, the estimated 8--10\% introgression rate is
much higher than Neanderthal gene flow, and would have plausibly been fertile
ground for dramatic selection for or against archaic-hominin-derived haplotypes\citep{Wall2019-ao}. By contrast, adaptation under a weakly
structured stem would have occurred continuously over much longer periods.
Patterns of polymorphism that are inconsistent with the single-stem model
predictions have been used to infer putative archaic admixed segments
\citep{Plagnol2006-lt,Hsieh2016-gk,Wall2019-ao,Durvasula2020-td}, negative
selection against such segments \citep{Wall2019-ao}, and pervasive positive
selection \citep{Schrider2017-kl}. However, such approaches are subject to high
false positives in the presence of population structure with migration
\citep{Petr2019-xo}, and their interpretation should be re-examined in light of
a weakly-structured-stem model within Africa.
